openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2021-03-16 paper/policar2020-opentsne.tex
pass through grammarly

This section is meant to familiarize the reader with the t-SNE algorithm and its
recent extensions. We motivate each extension and introduce standard notation.
This brief review is by no means comprehensive, but is self-contained and
introduces the mathematics upon which the \opentsne\ library is based.

@@

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method, which aims to find a low-dimensional
embedding, where local neighborhoods are preserved. More formally, given a
multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2,
\dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data
points in the data set, t-SNE aims to find a low dimensional embedding
$\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right
\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and
$\mathbf{x}_j$ are close in the high-dimensional space, their corresponding
embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is
primarily used as a visualization tool, $d$ is typically set to two. The
similarity between two data points in the high-dimensional space is defined as

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space. Perplexity is sometimes
interpreted as the continuous analog to the number of nearest neighbors to which
distances will attempt to be preserved.

@@

\noindent The objective is to find embeddings $\mathbf{Y}$ such that the
KL divergence is minimized.

The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.

@@

Belkina \textit{et al.} later found that faster convergence can be achieved by
increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a
side-effect, embeddings converge faster, and the number of iterations can be
lowered to 750, decreasing the overall runtime.  This convention has been
adopted by most modern t-SNE implementations.

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
impractical to any reasonably-sized data set and beckons for the development of
efficient approximation schemes. Van der Maaten observed that the t-SNE gradient
may be cast as an N-body problem where data points represent particles which
attract and repel each other~\cite{van2014accelerating}. The gradient from Eqn.
(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. This equation can then be viewed as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, which enable us
to greatly reduce the time complexity of t-SNE.

@@

Van der Maaten observed that evaluating the attractive forces between all pairs
of data points is excessive, and that considering only a handful of nearest
neighbors at each point is sufficient to obtain a good
approximation~\cite{van2014accelerating}.  Because distances are transformed to
similarities using a Gaussian kernel, and the bandwidth of each kernel is
selected such that only a predefined number of neighbors fall within the main
probability mass of the bell curve through the perplexity parameter, the
remaining data points fall into the exponentially decaying tails of each
Gaussian distribution.  These data points are assigned near-zero probabilities
and do not contribute to the overall attractive forces of data points.
Therefore it is sufficient to calculate the attractive forces for only a small
number of nearest neighbors instead of all $N$ points. By utilizing tree-based
nearest-neighbor search methods, the time complexity is thus reduced to
$\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realized that,
qualitatively, embeddings are visually indistinguishable when using only
\textit{approximate} nearest neighbors, further reducing time complexity to
$\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the
approach splits the space into quadrants and entire regions may be summarized by
simple statistics.  If a query point is far away from a given quadrant, the
repulsive forces exerted by all the points in that quadrant onto the query point
are summarized by a single point. This reduces the time complexity from
$\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes
$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
and FIt-SNE approximation schemes, reducing the time complexity to
$\mathcal{O}(M \log N)$ and $\mathcal{O}(\max \{ N, M \})$, respectively. In the
FIt-SNE approximation scheme, we additionally exploit the fact that the
reference embedding remains fixed throughout the optimization of newly added
points, and precompute the interpolation grid. This further reduces the runtime
complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
by selecting different perplexity values. Using a kernel with perplexity 500
captures long-range interactions which preserve global cluster organization.
Combining these with a more narrow kernel with perplexity 50 prevents small,
well-defined clusters from being absorbed into larger ones, leading a better
overall snapshot into the data structure.

@@

Standard t-SNE reveals the clustering structure at a single level of resolution.
Different perplexity parameter values can be used to identify global cluster
relationships or small, well-isolated groups. Unfortunately, varying perplexity
values can be time-consuming as this involves recomputing the $k$-nearest
neighbor graph which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.} suggest that varying the degree of freedom
in the t-distribution can be used to explore the clustering structure at
different levels of resolution~\cite{kobak2019heavy}.

@@

Embeddings produced by standard t-SNE often produce clusters separated by thin
boundaries and make use of all available space. While this is desirable at
times, it often obscures the global relationships between clusters as all
neighboring clusters appear about the same distance from one another. Other
dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where
clusters appear more compact and the white-space separating the clusters may be
interpreted at least partially as a loose measure of distance.

B{\"o}hm \textit{et al.} showed that the exaggeration factor $\rho$ can be used
to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
By incorporating exaggeration into later phases of optimization, t-SNE is able
introduce more white-space between clusters which better reflects the global
relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$
and $\rho=30$ produces embeddings visually similar to UMAP and ForceAtlas2,
respectively.

While it is difficult to claim one is better than the other, different settings
of $\rho$ may uncover different properties of the data manifold. For example, in
single-cell data, standard t-SNE often uncovers distinct, well-separated groups
of cell-types but obscures transitional paths between cell-states. This problem
is exacerbated when dealing with large numbers of data points. On the other
hand, ForceAtlas2 has been used successfuly to uncover trajectories and
transitions between cell types. Unfortunately, when using ForceAtlas2, large
clusters often absorb small, distinct groups of cells.  One might run t-SNE with
various parameter settings to avoid switching between different algorithms and
obtain similar visualizations.  We might generate an embedding using higher
levels of exaggeration to highlight the developmental transitions between cell
types and one with low levels of exaggeration to identify clear populations of
cells.

@@

t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout
algorithms, which operate on the $k$-nearest neighbor graph. Each method
constructs the graph differently and specifies the attractive and repulsive
forces in its own way, but ultimately, point positions are found by balancing
the attractive and repulsive forces between data points.

The final positions of data points in this class of algorithms is largely
dependent on the embedding initialization. UMAP found early success as it was
able to produce visualizations that better captured the global organization of
clusters. This was due to its initialization scheme, which initialized point
positions using Laplacian eigenmaps. On the other hand, most implementations of
t-SNE performed random initialization, which resulted in poor global coherence.
However, Kobak \& Linderman recently showed that using the same initialization
for both methods leads to visualizations, which exhibit similarly good global
coherency. Conversely, when initialized randomly, both methods produce
visualizations where clusters are arbitrarily positioned in the embedding space.

\opentsne\ defaults to using the two leading principal components as
initialization, but also provides a spectral approach similar to UMAP.  This
allows clusters in resulting embeddings to be organized in a more globally
coherent manner, leading to increased visualization interpretability.

>>>>

We here introduce notation, review the core t-SNE algorithm, and motivate and provide an overview of its recent extensions. The section presents the mathematics implemented in the \opentsne\ library.

@@

A t-distributed stochastic neighbor embedding (t-SNE) is a non-linear dimensionality reduction method that finds a low-dimensional embedding with preserved neighborhoods. More formally, given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization tool, $d$ is typically set to two. The similarity between two data points in the high-dimensional space is defined as

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying density of the data in the multi-dimensional space. We can view the perplexity as the continuous analog to the number of nearest neighbors to which the distances will be preserved.

@@

\noindent The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.

@@

Belkina \textit{et al.} later found that faster convergence can be achieved by increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a side-effect, embeddings converge faster, and the number of iterations can be
lowered to 750, decreasing the overall runtime.  Most modern t-SNE implementations have adopted this convention.

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes. Van der Maaten observed that the t-SNE gradient might be cast as an N-body problem where data points represent particles that attract and repel each other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us
to reduce the time complexity of t-SNE greatly.

@@

Van der Maaten observed that evaluating the attractive forces between all pairs
of data points is excessive, and that considering only a handful of nearest
neighbors at each point are sufficient to obtain a good
approximation~\cite{van2014accelerating}.  Because t-SNE transforms the distances to
similarities using a Gaussian kernel, and the bandwidth of each kernel is
selected such that only a predefined number of neighbors fall within the main
probability mass of the bell curve through the perplexity parameter, the
remaining data points fall into the exponentially decaying tails of each
Gaussian distribution.  These data points are assigned near-zero probabilities
and do not contribute to the overall attractive forces of data points.
Therefore it is sufficient to calculate the attractive forces for only a small
number of nearest neighbors instead of all $N$ points. By utilizing tree-based
nearest-neighbor search methods, the time complexity is thus reduced to
$\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realized that,
qualitatively, embeddings are visually indistinguishable when using only
\textit{approximate} nearest neighbors, further reducing time complexity to
$\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}), we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the
approach splits the space into quadrants, and simple statistics may summarize entire regions.  If a query point is far away from a given quadrant, the
repulsive forces exerted by all the points in that quadrant onto the query point
are summarized by a single point. This reduces the time complexity from
$\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes
$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
and FIt-SNE approximation schemes, reducing the time complexity to
$\mathcal{O}(M \log N)$ and $\mathcal{O}(\max \{ N, M \})$, respectively. In the
FIt-SNE approximation scheme, we additionally exploit the fact that the
reference embedding remains fixed throughout the optimization of newly added
points and precompute the interpolation grid. This further reduces the runtime
complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
by selecting different perplexity values. Using a kernel with perplexity 500
captures long-range interactions which preserve global cluster organization.
Combining these with a more narrow kernel with perplexity 50 prevents small,
well-defined clusters from being absorbed into larger ones, leading to an overall better
insight into the data structure.

@@

Standard t-SNE reveals the clustering structure at a single level of resolution.
We can use different perplexity parameter values to identify global cluster
relationships or small, well-isolated groups. Unfortunately, varying perplexity
values can be time-consuming as this involves recomputing the $k$-nearest neighbor graph, which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.} suggest that varying the degree of freedom
in the t-distribution can be used to explore the clustering structure at
different levels of resolution~\cite{kobak2019heavy}.

@@

Embeddings produced by standard t-SNE often have clusters separated by
thin boundaries and use all available space. While this is desirable,
it often obscures the global relationships between clusters as all
neighboring clusters appear about the same distance from one
another. Other dimensionality reduction methods such as
UMAP~\cite{2018arXivUMAP} or ForceAtlas2~\cite{jacomy2014forceatlas2}
tend to produce embeddings where clusters appear more compact and the
white-space separating the clusters may be interpreted at least
partially as a loose measure of distance.

B{\"o}hm \textit{et al.} showed that the exaggeration factor $\rho$
could be used to produce layouts more similar to UMAP and
ForceAtlas2~\cite{bohm2020unifying}. By incorporating exaggeration
into later phases of optimization, t-SNE introduces more white-space
between clusters which better reflects the global relations between
clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and
$\rho=30$ produces embeddings visually similar to UMAP and
ForceAtlas2, respectively.

While it is difficult to claim one is better than the other, different
$\rho$ may uncover different properties of the data manifold. For
example, standard t-SNE often exposes distinct, well-separated
cell-types in single-cell data but obscures transitional paths between
cell-states. This problem is exacerbated when dealing with large
numbers of data points. On the other hand, ForceAtlas2 has been used
successfully to uncover trajectories and transitions between cell
types. Unfortunately, when using ForceAtlas2, large clusters often
absorb small, distinct groups of cells. One might run t-SNE with
various parameter settings to avoid switching between different
algorithms and obtain similar visualizations. We might generate an
embedding using higher levels of exaggeration to highlight the
developmental transitions between cell types and one with low levels
of exaggeration to identify clear populations of cells.

@@

t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout
algorithms, which operate on the $k$-nearest neighbor graph. Each
method constructs the graph differently and specifies the attractive
and repulsive forces in its way. Still, ultimately, point positions
are found by balancing the attractive and repulsive forces between
data points.

The final positions of data points in this class of algorithms are
mainly dependent on the embedding initialization. UMAP found early
success as it produced visualizations that better captured the global
organization of clusters. This was due to its initialization scheme,
which initialized point positions using Laplacian eigenmaps. On the
other hand, most implementations of t-SNE performed random
initialization, which resulted in poor global coherence. However,
Kobak \& Linderman recently showed that using the same initialization
for both methods leads to visualizations, which exhibit similarly good
global coherency. Conversely, when initialized randomly, both methods
produce visualizations where clusters are arbitrarily positioned in
the embedding space.

\opentsne\ defaults to using the two leading principal components as
initialization, and besides, provides a spectral approach similar to
UMAP. This allows clusters in resulting embeddings to be organized in
a more globally coherent manner, leading to increased visualization
interpretability.
